---
title: "CleanedData"
output: html_document
date: "2024-04-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Packages + set up
load("opar.Rda")
library(readxl)
library(dplyr)
library(tidyverse)
library(tidyr)
library(readxl)
library(ggplot2)
library(maps)
library(sf)
library(geojsonsf)
library(rmapshaper)
library(magick)

library(algae)
source("funcs.R")
source("maps.R")
```

Subsets of Tasman/Marlborough

```{r}
bbox.tasmar <- c(1570,5370,1725,5520)*1000
xlim.tasmar <- bbox.tasmar[c(1,3)]
ylim.tasmar <- bbox.tasmar[c(2,4)]

bbox.marlselect <- c(1645,5415,1720,5495)*1000
bbox.marlselect <- c(1645,5412,1725,5502)*1000
xlim.marlselect <- bbox.marlselect[c(1,3)]
ylim.marlselect <- bbox.marlselect[c(2,4)]
xlim.havelock <- c(1660,1675)*1000
ylim.havelock <- c(5425,5440)*1000
xlim.nydia <- c(1662,1677)*1000
ylim.nydia <- c(5437,5450)*1000
xlim.arapawa <- c(1695,1718)*1000
ylim.arapawa <- c(5432,5452)*1000

# Coastal polygons for mapping
bbox.marlselect.polygon <- st_polygon(list(cbind(bbox.marlselect[c(1,3,3,1,1)],
                                                 bbox.marlselect[c(2,2,4,4,2)])))
coast.marlselect <- st_crop(coast, bbox.marlselect.polygon)

bbox.tasmar.polygon <- st_polygon(list(cbind(bbox.tasmar[c(1,3,3,1,1)],
                                             bbox.tasmar[c(2,2,4,4,2)])))
coast.tasmar <- st_crop(coast, bbox.tasmar.polygon)

# polygons in model
habspoly_sf <- geojson_sf("data/HABS_exp_6_release_polygons.geojson")
habspoly_sf <- st_transform(habspoly_sf, crs=st_crs(coast))
marlpoly_sf <- geojson_sf("data/Entire_marlb_polygons_volume_above_15m.geojson")
marlpoly_sf <- st_transform(marlpoly_sf, crs=st_crs(coast))
range(marlpoly_sf$CLUSTER_ID) # 1 409
if(max(marlpoly_sf$CLUSTER_ID)==409) {
  marlpoly_sf <- rename(marlpoly_sf, CLUSTER_ID.old=CLUSTER_ID)
  marlpoly_sf$CLUSTER_ID <- ifelse(marlpoly_sf$CLUSTER_ID.old>400,
                                   marlpoly_sf$CLUSTER_ID.old-1,marlpoly_sf$CLUSTER_ID)
}

marlpoly_sf.trim <- ms_erase(marlpoly_sf, coast.marlselect) 

globalcrs <- 4326
marlcrs <- st_crs(marlpoly_sf.trim)
```
```{r}
list.files("data/confidential")
```

## Locations

Note: corrected locations as follows in the file "Raw_MSQP_Locations-Corrected.csv"

```{r}
fname <- paste0("data/confidential/","Raw_MSQP_Locations-Corrected.csv")
habloc <- read.csv(fname)
# convert locations to sf object
habloc_sf <- st_as_sf(habloc %>% mutate(xp=x,yp=y), 
                      coords=c("xp","yp"),
                      crs=marlcrs,
                      agr="constant")

# identify which polygon each location falls into
habloc_sf <- habloc_sf %>% 
  mutate(intersection=as.integer(st_intersects(geometry, marlpoly_sf)),
         CLUSTER_ID=if_else(is.na(intersection),NA,marlpoly_sf$CLUSTER_ID[intersection])) %>%
  select(-intersection)

habloc <- habloc_sf %>% st_drop_geometry()
```

```{r}
plot(coast.tasmar$geometry, col="dark green", reset=FALSE, xlim=xlim.tasmar, ylim=ylim.tasmar)
rect(xlim.tasmar[1], ylim.tasmar[1], xlim.tasmar[2], ylim.tasmar[2])
plot(st_geometry(habloc_sf), pch=16, col="red", add=TRUE)
#points(habloc$x, habloc$y, pch=16, col="red")
title("Collection points in Tasman/Malborough")
```

```{r}
plot(coast.marlselect$geometry, col="dark green", xlim=xlim.marlselect, ylim=ylim.marlselect)
rect(xlim.marlselect[1], ylim.marlselect[1], xlim.marlselect[2], ylim.marlselect[2])
plot(st_geometry(habloc_sf), pch=16, col="red", add=TRUE)
#points(habloc$x, habloc$y, pch=16, col="red")
title("Collection points in Malborough")
```

Which points are not matched to a polygon?

```{r}
plot(coast.marlselect$geometry, col="dark green", xlim=xlim.marlselect, ylim=ylim.marlselect)
rect(xlim.marlselect[1], ylim.marlselect[1], xlim.marlselect[2], ylim.marlselect[2])
plot(st_geometry(habloc_sf %>% filter(is.na(CLUSTER_ID))), pch=16, col="red", add=TRUE)
#points(habloc$x, habloc$y, pch=16, col="red")
title("Collection points in Malborough\nnot matched to an ocean polygon")
```

```{r}
idx <- sapply(st_within(habloc_sf, bbox.marlselect.polygon),length)==1
#View(habloc_sf %>% filter(is.na(CLUSTER_ID) & idx)) #%>% as.data.frame()
outofarea_sf <- habloc_sf %>% filter(is.na(CLUSTER_ID) & idx) 

# Find the closest polygon
closest <- list()
for(i in seq_len(nrow(outofarea_sf))){
    closest[[i]] <- marlpoly_sf.trim[which.min(st_distance(marlpoly_sf.trim, outofarea_sf[i,])),]
}
outofarea_sf$CLUSTER_ID_closest <- sapply(closest, function(x) x$CLUSTER_ID)
outofarea <- outofarea_sf %>% st_drop_geometry()

outofarea_sf
## 6 sites, 2 repeats
```

 - G062 - D'Urville - fully outside tiled region - no match: remove
 - G086 - Head Nikau - close to shore - OK retain
 - G128 - Cook Strait 1 - too far out: remove
 - PG062 - D'Urville - fully outside tiled region - no match: remove
 - PG086 - Head Nikau - close to shore - OK retain
 - PG128 - Robertson Point - to far out to sea - no match: remove

```{r}
outofarea_retain_sf <- outofarea_sf %>% filter(siteID%in%c("G086","PG086"))
outofarea_retain_sf
```

```{r}
plot(st_geometry(coast.marlselect), reset=FALSE, col="darkgreen",
     main="Matching out of area sites to nearest polygon")
plot(st_geometry(marlpoly_sf.trim %>% filter(CLUSTER_ID%in%outofarea_retain_sf$CLUSTER_ID_closest)), col="yellow", add=TRUE)
plot(st_geometry(outofarea_retain_sf), pch=16, col="red", cex=1, add=TRUE)
axis(1); axis(2); box()
```

```{r}
habloc_sf <- merge(habloc_sf, st_drop_geometry(outofarea_retain_sf)[,c("siteID","CLUSTER_ID_closest")], all.x=TRUE)
idx <- is.na(habloc_sf$CLUSTER_ID) & !is.na(habloc_sf$CLUSTER_ID_closest)
habloc_sf$CLUSTER_ID[idx] <- habloc_sf$CLUSTER_ID_closest[idx]
habloc_sf$CLUSTER_ID_closest <- NULL
habloc <- habloc_sf %>% st_drop_geometry()
#write.csv(habloc, file="data/habloc.csv", row.names=FALSE)
```

```{r}
habloc <- read.csv("data/habloc.csv")
habloc_sf <- st_as_sf(habloc %>% mutate(xp=x,yp=y), 
                      coords=c("xp","yp"),
                      crs=marlcrs,
                      agr="constant")
```


## Detections

```{r, error=FALSE, message=FALSE, warning=FALSE}
fname <- paste0("data/confidential/","All MSQP phyto data to 01.05.23.xlsx")
habdat <- read_excel(fname) %>% as.data.frame()
#View(habdat)
nrow(habdat) # 292786
habdat <- habdat[habdat$Site_Code%in%habloc$siteID,]
nrow(habdat) # 101842
habdat[1:2,]

# Create an event ID: combination of site, date (differing times are minor)
habdat$date.time <- as.POSIXct(habdat$SAMPLED_DATE, format="%Y-%m-%d %H:%M:%S")
habdat$date <- as.Date(habdat$date.time)
habdat$year <- as.numeric(format(habdat$date,"%Y"))
habdat$event <- gsub(" ","-",paste(habdat$Site_Code, substring(habdat$SAMPLED_DATE,1,10), sep="-"))
# Retain only events from 2018 onwards
habdat <- habdat[habdat$year>=2018,]
nrow(habdat) # 52744
```

List of events
```{r}
event <- habdat %>% 
           group_by(event) %>%
           summarise(Site_Code=first(Site_Code), date=first(date), nsamples=n()) %>%
           ungroup()
nrow(event) # 7814
```



```{r}
species.names <- c("Alexandrium pacificum","Alexandrium catenella",
                   "Alexandrium spp.", 
                   "Alexandrium cf. pacificum","Alexandrium cf. catenella")
for(species in species.names) {
  etmp <- habdat %>%
    filter(REPORTED_NAME==species) %>%
    group_by(event) %>%
    summarise(n=n()) %>%
    ungroup()
  names(etmp) <- c("event",species)
  event[,species] <- NULL
  event <- merge(event,etmp,all.x=TRUE)
  event[,species][is.na(event[,species])] <- 0
}
```

```{r}
apply(event[,species.names],2,sum)
```

Remove any events where there is an cf. or spp. detection but no definite detection
```{r}
event$remove <- ((event$"Alexandrium pacificum"==0 & event$"Alexandrium catenella"==0)
                 &
                 (event$"Alexandrium spp.">0 
                  | event$"Alexandrium cf. pacificum">0 
                  | event$"Alexandrium cf. catenella">0)
                 )
sum(event$remove) # 58
```

Also remove the event on PG091 on 2014-02-26 at 13:53:00 - test detections, values unreliable
```{r}
event[event$event=="PG091-2014-02-26-13:53:00",]
event$remove[event$event=="PG091-2014-02-26-13:53:00"] <- TRUE
```

```{r}
nrow(habdat) # 52744
habdat <- habdat[!(habdat$event %in% event$event[event$remove]),]
nrow(habdat) # 52265
event <- event[!event$remove,]
nrow(event) # 7756
event$remove <- NULL
```

```{r}
etmp <- habdat %>%
    filter(REPORTED_NAME%in%species.names[1:2]) %>%
    group_by(event) %>%
    summarise(alex=n()) %>%
    ungroup()
event <- merge(event,etmp,all.x=TRUE)
event$alex[is.na(event$alex)] <- 0
nrow(event) # 7756
table(event$alex) # 754 detection events
```


```{r}
plot(event$date, event$alex)
```

Put each event in a polygon
```{r}
all(event$Site_Code %in% habloc$siteID) # TRUE
event <- merge(event, 
               habloc[,c("siteID","CLUSTER_ID","lon","lat","x","y")], 
               by.x="Site_Code", by.y="siteID", all.x=TRUE)
table(is.na(event$CLUSTER_ID)) # 1229 events NOT inside polygons 6527 events are in polygons
#table(event$Site_Code, is.na(event$CLUSTER_ID))
plot(event$x, event$y, col=ifelse(is.na(event$CLUSTER_ID),"blue","red"))
event <- event[!is.na(event$CLUSTER_ID),]
nrow(event) # 6527 events
habdat <- habdat[habdat$event %in% event$event,]
nrow(habdat) # 45346
```

Assign a week number to each event
```{r}
event$week <- date.as.week(event$date)
event$year <- as.numeric(format(event$date,"%Y"))
table(event$week)
#any(is.na(event$week))
sum(event$alex) # 725 events with a detection
```

Check for multiple events at the same location in the same week
```{r}
tt <- table(paste(event$Site_Code,event$week))
tt[tt>1]
# quite a few instances of this
sum(tt[tt>1]) # 310 events
length(tt[tt>1]) # 155 occasions
ee <- names(tt[tt>1])
event$alex[paste(event$Site_Code,event$week)%in%ee]
View(event[paste(event$Site_Code,event$week)%in%ee,] %>% arrange(year,event))
```

For simplicity,  identify the first event in each week with a positive detection
```{r}
fevent <- event %>%
  group_by(Site_Code,week) %>%
  arrange(desc(alex),date) %>%
  summarise(fevent=first(event),nobs=n(),alex=sum(alex)) %>%
  ungroup()
nrow(event) # 6527
nrow(fevent) # 6372
event$include <- event$event %in% fevent$fevent
sum(event$include) # 6372
sum(!event$include) # 155
sum(event$alex[!event$include]) # removes 8 detections of Alexandrium
```

```{r}
alex <- habdat[habdat$REPORTED_NAME %in% species.names[1:2],]
is.unique(alex$event) # TRUE
nrow(alex) # 725
```

```{r}
nrow(event) # 6527
event <- merge(event, alex[,c("event","REPORTED_NAME","Data_Value")],all.x=TRUE)
nrow(event) # 6527
```

All units are cells/litre

```{r}
table(alex$Data_Value)
event$Value <- ifelse(is.na(event$Data_Value) | 
                        event$Data_Value=="Not Detected",0,
                      as.numeric(event$Data_Value))
table(event$Value,exclude=NULL)
```

Save the results

```{r}
event <- event %>%
  select(event,include,Site_Code,date,week,year,
         nsamples,alex,CLUSTER_ID,Value,
         lon,lat,x,y,REPORTED_NAME,Data_Value)
#write.csv(event,"data/events_alex.csv",row.names=FALSE)
```

```{r}
event <- read.csv("data/events_alex.csv")
nrow(event) # 6527
event <- event[event$include,]
nrow(event) # 6372
event$date <- as.Date(event$date)
```


