---
title: "eDNA"
author: "Richard Arnold"
date: '2024-07-16'
output:
  pdf_document: default
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Detection of eDNA

 * A volume $V_S$ of water is collected, and all of the eDNA it contains is collected by filtration.
   Assume a count $N_j$ of eDNA fragments of type $j$ is collected ($j=1,\ldots,J$). 
   If the number density of fragments in the water is $\nu_j$ then
   \[
      N_j | \nu_j, V_S \sim \text{Poisson}(\nu_jV_S)
   \]
 * The collected eDNA is mixed with a volume $V_M$ of a proprietary mix of primers, probes and water;
 * A test volume $V_T$ ($1\mu\ell$) is taken from the mixture: The count of fragments in the test volume is
   \[
      n_j | N_j,V_M,V_T\sim\text{Poisson}(N_jV_T/V_M)
   \]
 * The test volume is atomised into $m$ droplets.  The count of fragments in droplet $i$ is 
   \[
      y_{ji} | n_j,m \sim\text{Poisson}(n_j/m)
   \]
 * The probability that droplet $i$ has zero fragments is
   \[
      \Pr(y_{ji}=0 | n_j,m) = \exp\left(-\frac{n_j}{m}\right)
   \]
   and thus the probability that it contains any fragments is
   \[
      \Pr(y_{ji}>0 | n_j,m) = 1-\exp\left(-\frac{n_j}{m}\right)
   \]
 * The eDNA in the droplets is amplified by PCR, and $m_{j+}$ of the droplets are observed to 
   be positive for fragment type $j$.  Thus the observed proportion of positive 
   droplets is 
   \[
      \widehat{p}_+ = \frac{m_{j+}}{m}
   \]
   and
   \[
      m_{j+} | n_j,m \sim \text{Binomial}\left(m,1-e^{-n_j/m}\right)
   \]
   or if $n_j/m$ is very small then
   \[
      m_{j+} | n_j,m \sim \text{Poisson}\left(m\left(1-e^{-n_j/m}\right)\right)
   \]
   Note that we assume no false positives and no false negatives when individual 
   droplets are amplified and tested for the presence of the eDNA.  

This process is aimed at determining $\nu_j$, the original number density in the water
from which the sample was collected.  A simple estimate of $\mu_j$ is 
\[
   \widehat{\nu}_j = -\frac{V_M m}{V_SV_T} \log\left(1-\frac{m_{j+}}{m}\right)
\]
However in this application we are most interested in the probability that the process leads to negative result
when $\nu_j$ is non-zero. i.e. we want to know
\begin{eqnarray}
  && \mbox{}\hspace{-2cm}\Pr(m_{j+}=0 | m,V_S,V_T,V_M,\nu_j)\\
  &=& 
  \sum_{N_j=0}^\infty \sum_{n_j=0}^\infty
  \Pr(m_{j+}=0 | n_j,m) \Pr(n_j | N_j,V_M,V_T) \Pr(N_j | \nu_j,V_S)\\
  &=& 
  \sum_{N_j=0}^\infty \sum_{n_j=0}^\infty
  \exp\left(-m\left[1-e^{-n_j/m}\right]\right) 
  \exp\left(-\frac{N_jV_T}{V_M}\right)\frac{\left(\frac{N_jV_T}{V_M}\right)^{n_j}}{n_j!}
  \exp\left(-\nu_jV_S\right)\frac{\left(\nu_jV_S\right)^{N_j}}{N_j!}
\end{eqnarray}
Note that the sum over $n_j$ should only be from 0 to $N_j$, but we sum to infinity because the 
additional terms are all negligible.  

Writing the ratio $V_T/V_M$, i.e. the proportion of the mixture that is atomised and tested, as $d$,
and suppressing the index $j$
\begin{eqnarray}
  && \mbox{}\hspace{-2cm}\Pr(m_{+}=0 | m,V_S,d,\nu)\\
  &=& 
  \sum_{N=0}^\infty \sum_{n=0}^\infty
  \Pr(m_{+}=0 | n,m) \Pr(n | N,d) \Pr(N | \nu,V_S)\\
  &=& 
  \sum_{N=0}^\infty \sum_{n=0}^\infty
  \exp\left(-m\left[1-e^{-n/m}\right]\right) 
  e^{-Nd}\frac{(Nd)^{n}}{n!}
  e^{-\nu V_S}\frac{(\nu V_S)^{N}}{N!}
\end{eqnarray}

At very low concentrations
\[
   n | \nu, V_S, d \sim \text{Poisson}(\nu V_Sd)
\]
so that
\begin{eqnarray}
  && \mbox{}\hspace{-2cm}\Pr(m_{+}=0 | m,V_S,d,\nu)\\
  &=& 
  \sum_{n=0}^\infty
  \Pr(m_{+}=0 | n,m) \Pr(n | \nu,V_S,d)\\
  &=& 
  \sum_{n=0}^\infty
  \exp\left(-m\left[1-e^{-n/m}\right]\right) 
  e^{-\nu V_S d}\frac{(\nu V_S d)^{n}}{n!}\\
  &=& 
  \sum_{n=0}^\infty
  \exp\left(-m\left[1-e^{-n/m}\right]\right) 
  e^{-\lambda}\frac{\lambda^{n}}{n!}\\
\end{eqnarray}
where $\lambda = \nu V_S d$.

Typically we have $m=2\times10^4$ droplets.  We might expect $\nu V_s\simeq 10$ fragments to be catpured, 
and a subsampling of $d=1/20$, and in that case $\lambda\simeq 0.5$

```{r}
probmp0 <- function(lambda, m=20000, nbig=20000) {
  nvec <- 0:nbig
  fval <- sapply(lambda, function(lambdav) sum(exp(-m*(1-exp(-nvec/m)))*dpois(nvec,lambdav)))
  return(fval)
}
m <- 20000; nbig <- m
np <- 101
lambdavec <- 10^(seq(from=-3,to=+1,length=np))
plot(lambdavec, (probmp0(lambdavec, m=m, nbig=nbig)), type="l",
     xlab=expression(lambda), ylab=expression("Pr("*m["+"]==0*"): Prob of no detections"), 
     main=bquote(lambda==nu*V[S]*d*";  "*m*"="*.(m)*" droplets"))
```


## Particle dispersal

If there is a mass released at time $t_0$ at location ${\bf x}_0$ then it is transported
by advection (water flow) and dispersion (turbulence).  

Let $p({\bf x},t | {\bf x}_0, t_0, {\bf v}(), {\bf a}(), \Psi)$ be the proportion of the mass released
that is found at location ${\bf x}$ at time $t$, given water flow ${\bf v}()$ and dispersion ${\bf a}()$.
$\Psi$ is a set of additional parameters controlling the flow/dispersion model.

A simple model is a simple Gaussian spread in 2D:
\begin{eqnarray}
   p({\bf x},t | {\bf x}_0, t_0, {\bf v}, {\bf a}, \Psi)
   &=& 
   \frac{I(t-t_0\geq 0)}{2\pi \sqrt{a_xa_y}(t-t_0+t_\varepsilon)}
   \exp\left(
      -\frac{\left(x-x_0-v_x(t-t_0)\right)^2}{2a_x(t-t_0+t_\varepsilon)}
      -\frac{\left(y-y_0-v_y(t-t_0)\right)^2}{2a_y(t-t_0+t_\varepsilon)}
       \right)\\
   &=& g({\bf x},t-t_0 |{\bf x}_0, {\bf v}, {\bf a}, \Psi)
\end{eqnarray}
where a small positive temporal offset $\Psi=\{t_\varepsilon\}$ is added to avoid a singularity at $t=t_0$.

We have 
\[
   \int\!\!\int p({\bf x},t|{\bf x}_0,t_0, {\bf v},{\bf a}, \Psi) \,{\rm d}{\bf x}= 1 
\]
for all times $t>t_0$.

Note that $g()$ has a symmetry in which we can interchange ${\bf x}$ and ${\bf x}_0$ and at the
same time reverse the velocity field, so for any time interval $u\geq 0$:
\begin{equation}
   g({\bf x},u|{\bf x}_0,{\bf v},{\bf a},\Psi)
   = 
   g({\bf x}_0,u|{\bf x},-{\bf v},{\bf a},\Psi)
\end{equation}

```{r class.source = 'fold-show'}
source("funcs.R")
```

```{r}
nx <- 101
ny <- 101
xmin <- 0; xmax <- 100
ymin <- 0; ymax <- 100
xvec <- seq(from=xmin, to=xmax, length=nx)
yvec <- seq(from=ymin, to=ymax, length=ny)

x0 <- mean(c(xmin,xmax))
y0 <- mean(c(ymin,ymax))
t0 <- 0
vx <- 5; vy <- 1
ax <- 2^2; ay <- 4^2
teps <- 0.01
```

For example here is the spread over time of mass released at ${\bf x}_0=(`r x0`,`r y0`)^T$
at time $t=`r t0`$ with constant flow field ${\bf v}=(`r vx`,`r vy`)^T$ and
dispersion ${\bf a}=(`r ax`,`r ay`)^T$ and with $t_\varepsilon=`r teps`$.

**Forward tracking:** $p({\bf x},t|{\bf x}_0,t_0, {\bf v}(), {\bf a}(), \Psi)$: 
$t_0, \ldots, t_0+10$ 
(no decay)
```{r}
par(mfrow=c(3,4))
par(mar=0.1*c(1,1,1,1))
colvec <- hcl.colors(12, "YlOrRd", rev = TRUE)
colvec <- hcl.colors(12, "YlGnBu")#, rev = TRUE)
for(t in 0:10) {
   amat <- outer(xvec,yvec, dispfunc, t=t,
                 x0=x0,y0=y0,t0=0, vx=vx,vy=vy, ax=ax,ay=ay, teps=teps)
   #if(t==0) amax <- max(amat)
   amax <- max(amat)
   breaks <- seq(from=0, to=amax, length=length(colvec)+1)
   image(xvec, yvec, amat, asp=1, col=colvec, breaks=breaks, axes=FALSE)
   #image(xvec, yvec, amat,asp=1, main=bquote("Forwards:" ~ t==.(t)), xlab="x", ylab="y",
   #      col=colvec, breaks=breaks)
}
```

A simple backtracking model simply reverses the flow field.  If we are interested in the likely origin
of mass observed at location ${\bf x}_1$ at time $t_1$ we can backtrack by reversing ${\bf v}()$.  

We might assume that the likelihood that the observed mass was released at location ${\bf x}$ at time $t<t_1$ 
is given by
\begin{eqnarray}
   \tilde{p}({\bf x},t | {\bf x}_1,t_1, {\bf v}(), {\bf a}(), \Psi)
     &=& g({\bf x},t_1-t |{\bf x}_1, -{\bf v}(), {\bf a}(), \Psi)\\
     &=& p({\bf x},-t | {\bf x}_1,-t_1, -{\bf v}(), {\bf a}(), \Psi)
\end{eqnarray}
In the simple Gaussian model above this is
\begin{eqnarray}
   &&\mbox{}\hspace{-2cm}\tilde{p}({\bf x},t | {\bf x}_1, t_1, {\bf v}(), {\bf a}(), \Psi)\\
   &=& 
   p({\bf x},t_1 | {\bf x}_1, t, -{\bf v}(), {\bf a}(), \Psi)\\
   &=& 
   p({\bf x},-t | {\bf x}_1, -t_1, -{\bf v}(), {\bf a}(), \Psi)\\
   &=& 
   g({\bf x},t_1-t | {\bf x}_1, -{\bf v}(), {\bf a}(), \Psi)\\
   &=& 
   \frac{I(t_1-t\geq 0)}{2\pi\sqrt{a_xa_y}(t_1-t+t_\varepsilon)}
   \exp\left(
      -\frac{\left(x-x_1+v_x(t_1-t)\right)^2}{2a_x(t_1-t+t_\varepsilon)}
      -\frac{\left(y-y_1+v_y(t_1-t)\right)^2}{2a_y(t_1-t+t_\varepsilon)}
       \right)
\end{eqnarray}

**Back tracking:** $\tilde{p}({\bf x},t|{\bf x}_1,t_1, {\bf v}(), {\bf a}(), \Psi) = p({\bf x},-t|{\bf x}_1,-t_1, -{\bf v}(), {\bf a}(), \Psi)$: $t_1, \ldots, t_1-10$ 
(no decay)
```{r}
par(mfrow=c(3,4))
par(mar=0.1*c(1,1,1,1))
x1 <- 100; y1 <- 60
t1 <- 10
for(t in 10:0) {
   amat <- outer(xvec,yvec, dispfunc, t=-t,
                 x0=x1,y0=y1,t0=-t1, vx=-vx,vy=-vy, ax=ax,ay=ay, teps=teps)
   #if(t==10) amax <- max(amat)
   amax <- max(amat)
   breaks <- seq(from=0, to=amax, length=length(colvec)+1)
   image(xvec, yvec, amat,asp=1, col=colvec, breaks=breaks, axes=FALSE)
   #image(xvec, yvec, amat,asp=1, main=bquote("Backwards: " ~ t==.(t)), xlab="x", ylab="y",
   #      col=colvec, breaks=breaks)
}
```

## Continuous release with decay

Assume that at location ${\bf x}_0$ there is a continuous release of mass
at a rate $N_0$ kg/s.   Further assume that the mass decays at a rate
\[
   d(t|T_h,T_m) = 2^{-t/T_h} I(0\leq t\leq T_m)
\]
where $T_h$ is the decay half life and $T_m$ is the maximum age at which
all particles have disintegrated.  If we set $\lambda = (1/T_h)\log 2$ then
\[
   d(t|T_h,T_m) = e^{-\lambda t} I(0\leq t\leq T_m)
\]
The steady state mass in the system is then
\begin{eqnarray}
  M_0 &=& \int_{-\infty}^t N_0 d(t-t_0|T_h,T_m)\;{\rm d}t_0\\
      &=& N_0 \int_{-\infty}^t e^{-\lambda(t-t_0)} I(t_0>t-T_m)\;{\rm d}t_0\\
      &=& N_0 \int_{t-T_m}^t e^{-\lambda(t-t_0)} \;{\rm d}t_0\\
      &=& N_0 \int_0^{T_m} e^{-\lambda u} \;{\rm d}u\\
      &=& \frac{N_0}{\lambda}\left[1-e^{-\lambda T_m}\right]\\
      &=&\frac{N_0T_h}{\log 2}\left[1-e^{-\lambda T_m}\right]
\end{eqnarray}

The mass density transiting per unit time at location ${\bf x}$ at time $t$ 
emitted from a point source at ${\bf x}_0$ at time $t_0$ is
\[
   \frac{{\rm d}\rho({\bf x},t|N_0,{\bf x}_0,t_0,{\bf v}(),{\bf a}(), \Psi)}{{\rm d}t}
   = 
   N_0 d(t-t_0|T_h,T_m) p({\bf x},t|{\bf x}_0,t_0,{\bf v}(),{\bf a}(),\Psi)
\]

**Density rate** ${\rm d}\rho({\bf x},t|N_0,{\bf x}_0,t_0,{\bf v}(),{\bf a}(), \Psi)/{\rm d}t$, 
$t=t_0, \ldots, t_0+10$ (with decay).  (Scales differ between panels: note max and sum values.)
```{r}
par(mfrow=c(3,4))
par(mar=0.1*c(1,1,1,1))
n0 <- 1
th <- 5
tm <- 7
for(t in 0:10) {
   amat <- outer(xvec,yvec, denratefunc, t=t,
                 x0=x0,y0=y0,t0=0, vx=vx,vy=vy, ax=ax,ay=ay, teps=teps, n0=n0,th=th,tm=tm)
   #amat <- log(1+amat)
   #if(t==0) {
   #   amin <- min(amat)
   #   amax <- max(amat)
   #}
   amin <- min(amat)
   amax <- max(amat)
   breaks <- seq(from=amin, to=amax, length=length(colvec)+1)
   image(xvec, yvec, amat, asp=1, col=colvec, breaks=breaks, axes=FALSE)
   #image(xvec, yvec, amat,asp=1, main=bquote("Forwards with decay:" ~ t==.(t)), xlab="x", ylab="y",
   #      col=colvec, breaks=breaks)
   mtext(bquote("sum:"~.(sum(amat))~"max:"~.(max(amat))), side=3, adj=1, cex=0.6, line=-1, col="white")
}
```

The total (equilibrium) mass density present is
\begin{eqnarray}
   &&\mbox{}\hspace{-2cm}\rho({\bf x},t|N_0,{\bf x}_0,{\bf v}(),{\bf a}(),\Psi)\\
   &=& \rho({\bf x}|N_0,{\bf x}_0,{\bf v}(),{\bf a}(),\Psi)\\
   &=& \int_{-\infty}^t N_0 d(t-t_0|T_h,T_m) p({\bf x},t|{\bf x}_0,t_0,{\bf v}(),{\bf a}(),\Psi)\;{\rm d}t_0\\
   &=& N_0 \int_{t-T_m}^t e^{-\lambda (t-t_0)} g({\bf x},t-t_0|{\bf x}_0,{\bf v}(),{\bf a}(),\Psi)\;{\rm d}t_0\\
   &=& N_0 \int_0^{T_m} e^{-\lambda u} g({\bf x},u|{\bf x}_0,{\bf v}(),{\bf a}(),\Psi)\;{\rm d}u\\
   &=& N_0 h({\bf x}|{\bf x}_0,{\bf v}(),{\bf a}(),\Psi)
\end{eqnarray}
where
\[
  h({\bf x}|{\bf x}_0,{\bf v}(),{\bf a}(),\Psi)
  =
  \int_0^{T_m} e^{-\lambda u} g({\bf x},u|{\bf x}_0,{\bf v}(),{\bf a}(),\Psi)\;{\rm d}u
\]
The function $h()$ retains the same symmetry as $g()$, so since
\[
  g({\bf x},u|{\bf x}_0,{\bf v},{\bf a},\Psi)
  = 
  g({\bf x}_0,u|{\bf x},-{\bf v},{\bf a},\Psi)
\]
it follows that 
\[
  h({\bf x},u|{\bf x}_0,{\bf v},{\bf a},\Psi)
  = 
  h({\bf x}_0,u|{\bf x},-{\bf v},{\bf a},\Psi)
\]

In the Gaussian dispersal example
\[
  h({\bf x}|{\bf x}_0,{\bf v}(),{\bf a}(),\Psi)
  =
  \frac{1}{2\pi \sqrt{a_xa_y}} \int_0^{T_m} \frac{e^{-\lambda u}}{u+t_\varepsilon}
      \exp\left(
         -\frac{\left(x-x_0-v_xu\right)^2}{2a_x(u+t_\varepsilon)}
         -\frac{\left(y-y_0-v_yu\right)^2}{2a_y(u+,t_\varepsilon)}       
         \right)
\]

```{r}
invisible()
#th <- 5
#tm <- 7
#ax <- 2^2; ay <- 5^2
#amat <- array(0,dim=c(nx,ny))
#nc <- 100
#for(i in 1:nc) {
#   t <- i/nc * 10
#   amat <- amat + outer(xvec,yvec, denratefunc, t=t,
#                        x0=x0,y0=y0,t0=0, vx=vx,vy=vy, ax=ax,ay=ay, teps=teps, n0=n0,th=th,tm=tm)
#}
#image(xvec, yvec, amat,asp=1, 
#      main=bquote(t[h]==.(th) ~ ", " ~ t[m]==.(tm)), 
#      xlab="x", ylab="y",col=colvec)
invisible()
```

**Equilibrium Density**: 
$\rho({\bf x}|N_0,{\bf x}_0,{\bf v},{\bf a},\Psi)=N_0h({\bf x}|{\bf x}_0,{\bf v},{\bf a},\Psi)$
```{r}
th <- 5
tm <- 7
ax <- 2^2; ay <- 5^2
amat <- n0*outer(xvec,yvec, hfunc,
                 x0=x0,y0=y0, vx=vx,vy=vy, ax=ax,ay=ay, teps=teps, th=th,tm=tm)
image(xvec, yvec, amat, asp=1, 
      main=bquote(h(x*"|"*x[0])*":" ~ x[0]==(.(x0)*","*.(y0)) ~ "," ~ t[h]==.(th) ~ "," ~ t[m]==.(tm)), xlab="x", ylab="y",col=colvec)
```

The above figure plots, for a continuous fixed source of rate $N_0$ at location ${\bf x}_0$, 
the equilibrium density at all locations ${\bf x}$: $N_0 h({\bf x}|{\bf x}_0)$.

Now plot at each ${\bf x}_0$ the equilibrium density at ${\bf x}_1$ (in the centre of the diagram)
that would result from a continuous source at ${\bf x}_0$.  i.e. $N_0 h({\bf x}_1|{\bf x_0})$
\begin{eqnarray*}
   N_0 \tilde{h}({\bf x}_0 | {\bf x}_1, {\bf v}, {\bf a}, \Psi) 
    &=& N_0 h({\bf x}_1|{\bf x}_0, {\bf v}, {\bf a}, \Psi)\\
    &=& N_0 h({\bf x}_0|{\bf x}_1, -{\bf v}, {\bf a}, \Psi)
\end{eqnarray*}


$N_0 \tilde{h}({\bf x}_0|{\bf x}_!, {\bf v}, {\bf a}, \Psi) 
  = N_0 h({\bf x}_1|{\bf x}_0, {\bf v}, {\bf a}, \Psi) 
  = N_0 h({\bf x}_0|{\bf x}_1, -{\bf v}, {\bf a}, \Psi)$
```{r}
th <- 5
tm <- 7
ax <- 2^2; ay <- 5^2
x1 <- 50; y1 <- 50
rho1 <- 1; mu <- 1
amat <- outer(xvec,yvec, function(x0vec,y0vec) {
                    apply(cbind(x0vec,y0vec),1,function(x0y0) {
                          hfunc(x=x1,y=y1,x0=x0y0[1],y0=x0y0[2],
                          vx=vx,vy=vy, ax=ax,ay=ay, teps=teps, th=th,tm=tm)
                    })
                 })
dim(amat) <- c(length(xvec),length(yvec))
image(xvec, yvec, amat, asp=1, 
      main=bquote(h(x[1]*"|"*x[0])*":" ~ x[1]==(.(x0)*","*.(y0)) ~ "," ~ t[h]==.(th) ~ "," ~ t[m]==.(tm)), 
      xlab=expression(x[0]), ylab=expression(y[0]), col=colvec)
```

(Computing this the other way using the symmetry of h())
```{r}
th <- 5
tm <- 7
ax <- 2^2; ay <- 5^2
x1 <- 50; y1 <- 50
rho1 <- 1; mu <- 1
amat <- n0*outer(xvec,yvec, hfunc,
                 x0=x1,y0=y1, vx=-vx,vy=-vy, ax=ax,ay=ay, teps=teps, th=th,tm=tm)
image(xvec, yvec, amat, asp=1, 
      main=bquote(h(x[1]*"|"*x[0])*":" ~ x[1]==(.(x0)*","*.(y0)) ~ "," ~ t[h]==.(th) ~ "," ~ t[m]==.(tm)), 
      xlab=expression(x[0]), ylab=expression(y[0]), col=colvec)
```

## Probabilistic Inversion

Now assume that at some location ${\bf x}_1$ we observe a density $\rho_1$, and that 
this in an observation made with error:
\[
   \rho_1 | \rho_1^\ast, \sigma_e \sim N(\rho_1^\ast, \sigma_e^2)
\]
were the true density is $\rho_1^\ast$ and the error variance is $\sigma_e^2$.

Assume that there is a single continous source of unknown strength $N_0$ and unknown
location ${\bf x}_0$.   Priors for the strength and location are $\pi(N_0)$ and $\pi({\bf x}_0)$
respectively.

Assume that the velocity field ${\bf v}()$ and the diffusion field ${\bf a}()$ are known.

Using the results from the previous section, conditional on $N_0$ and ${\bf x}_0$ the 
observed density at ${\bf x}_1$ is
\[
  \rho_1^\ast= N_0 h({\bf x}_1|{\bf x}_0, {\bf v}(), {\bf a}(), \Psi)
\]
which we write for brevity as 
\[
   \rho_1^\ast = h_1^\ast({\bf x}_0) = h_1^\ast
\]
but noting its dependence on ${\bf x}_0$.

It follows that the joint distribution of the unknown source strength $N_0$ and location ${\bf x}_0$ is
\begin{eqnarray*}
   p(N_0,{\bf x}_0|\rho_1,{\bf x}_1)
   &\propto&
   \pi(N_0) \pi({\bf x}_0) \exp\left(-\frac{1}{2\sigma_e^2}
           \left[\rho_1-N_0h_1^\ast({\bf x}_0)\right]^2\right)\\
   &\propto&
   \pi(N_0) \pi({\bf x}_0) \exp\left(-\frac{(h_1^\ast)^2}{2\sigma_e^2}
           \left[N_0-\frac{\rho_1}{h_1^\ast}\right]^2\right)
\end{eqnarray*}
A suitable prior for $N_0$ is the Exponential($\mu$) distribution
\begin{eqnarray*}
  N_0  &\sim& \text{Exp}(\mu)\\
  \pi(N_0) &=& \mu e^{-\mu N_0}
\end{eqnarray*}
This can be marginalised out of the posterior distribution as follows. 
Firstly the joint distribution is
\begin{eqnarray*}
   p(N_0,{\bf x}_0|\rho_1,{\bf x}_1)
   &\propto&
   \pi({\bf x}_0) e^{-\mu N_0} 
           \exp\left(-\frac{(h_1^\ast)^2}{2\sigma_e^2}
           \left[N_0-\frac{\rho_1}{h_1^\ast}\right]^2\right)\\
   &\propto& 
   \pi({\bf x}_0)
           \exp\left( -\frac{(h_1^\ast)^2}{2\sigma_e^2}
           \left[N_0-\left(\frac{\rho_1}{h_1^\ast} -
                           \frac{\mu\sigma_e^2}{(h_1^\ast)^2}
                           \right)\right]^2
           - \frac{\rho_1\mu}{h_1^\ast} + \frac{\mu^2\sigma_e^2}{2(h_1^\ast)^2}
                           \right)
\end{eqnarray*}
Integrating out $N_0$ leads to
\begin{eqnarray*}
   p({\bf x}_0|\rho_1,{\bf x}_1)
   &\propto&
   \pi({\bf x}_0) \frac{\sigma_e}{h_1^\ast}
     \exp\left(
           - \frac{\rho_1\mu}{h_1^\ast} + \frac{\mu^2\sigma_e^2}{2(h_1^\ast)^2}
                           \right)
\end{eqnarray*}
If we let $\sigma_e\rightarrow 0$ (indicating very accurate observations) we
are left with a very simple form for the posterior:
\begin{eqnarray*}
   p({\bf x}_0|\rho_1,{\bf x}_1)
   &\propto&
   \pi({\bf x}_0) \frac{1}{h_1^\ast} e^{\frac{\rho_1\mu}{h_1^\ast}}
\end{eqnarray*}
which we plot below for a uniform prior on the source location $\pi({\bf x}_0)\propto 1$.

```{r}
th <- 5
tm <- 7
ax <- 2^2; ay <- 5^2
x1 <- 50; y1 <- 50
rho1 <- 1; mu <- 1
amat <- n0*outer(xvec,yvec, hfunc,
                 x0=x1,y0=y1, vx=-vx,vy=-vy, ax=ax,ay=ay, teps=teps, th=th,tm=tm)

qmat <- exp(-max(amat)/3.0/amat) * (1/amat)
image(xvec, yvec, qmat, asp=1, 
      main=bquote(p(x[0]*"|"*x[1])*":" ~ x[1]==(.(x0)*","*.(y0)) ~ "," ~ t[h]==.(th) ~ "," ~ t[m]==.(tm)), 
      xlab=expression(x[0]), ylab=expression(y[0]), col=colvec)
```

Plotting again, but as a contour map
```{r}
par(mfrow=c(1,2))
contour(amat, main=expression(h(x[0]*"|"*x[1])))
contour(qmat, main=expression(p(x[0]*"|"*x[1])))
```



A different prior: e.g. normally distributed in $y$ (centered on $y_a$ with 
standard deviation $\sigma_a$) and uniform in $x$:
\[
   \pi({\bf x}_0) \propto \exp\left(-\frac{1}{2\sigma_a^2}\left[x_{0y}-y_a\right]^2\right)
\]


```{r}
ya <- 60
sigmaa <- 10
p0mat <- t(array(exp(-0.5*(yvec-ya)^2/sigmaa^2), dim=dim(amat)))
image(xvec, yvec, p0mat, main=bquote(pi(x[0]) %prop% N[y](.(ya),.(sigmaa)^2)), 
      xlab=expression(x[0]), ylab=expression(y[0]))
```

The effect of the prior is shown at right below
```{r}
par(mfrow=c(1,2))
image(xvec, yvec, qmat, asp=1, 
      main=bquote(p(x[0]*"|"*x[1])*":"  ~ pi(x[0]) %prop% 1), 
      xlab=expression(x[0]), ylab=expression(y[0]), col=colvec)
image(xvec, yvec, p0mat*qmat, asp=1, 
      main=bquote(p(x[0]*"|"*x[1])*":" ~ pi(x[0]) %prop% N[y](.(ya),.(sigmaa)^2)), 
      xlab=expression(x[0]), ylab=expression(y[0]), col=colvec)
```

