---
title: "habanalysis"
output: html_document
date: "2024-04-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Packages + set up
load("opar.Rda")
library(readxl)
library(dplyr)
library(tidyverse)
library(tidyr)
library(readxl)
library(ggplot2)
library(maps)
library(sf)
library(geojsonsf)
library(rmapshaper)
library(magick)

library(algae)
source("funcs.R")
source("maps.R")
```

Subsets of Tasman/Marlborough

```{r}
bbox.tasmar <- c(1570,5370,1725,5520)*1000
xlim.tasmar <- bbox.tasmar[c(1,3)]
ylim.tasmar <- bbox.tasmar[c(2,4)]

bbox.marlselect <- c(1645,5415,1720,5495)*1000
bbox.marlselect <- c(1645,5412,1725,5502)*1000
xlim.marlselect <- bbox.marlselect[c(1,3)]
ylim.marlselect <- bbox.marlselect[c(2,4)]
xlim.havelock <- c(1660,1675)*1000
ylim.havelock <- c(5425,5440)*1000
xlim.nydia <- c(1662,1677)*1000
ylim.nydia <- c(5437,5450)*1000
xlim.arapawa <- c(1695,1718)*1000
ylim.arapawa <- c(5432,5452)*1000

# Coastal polygons for mapping
bbox.marlselect.polygon <- st_polygon(list(cbind(bbox.marlselect[c(1,3,3,1,1)],
                                                 bbox.marlselect[c(2,2,4,4,2)])))
coast.marlselect <- st_crop(coast, bbox.marlselect.polygon)

bbox.tasmar.polygon <- st_polygon(list(cbind(bbox.tasmar[c(1,3,3,1,1)],
                                             bbox.tasmar[c(2,2,4,4,2)])))
coast.tasmar <- st_crop(coast, bbox.tasmar.polygon)

# polygons in model
habspoly_sf <- geojson_sf("data/HABS_exp_6_release_polygons.geojson")
habspoly_sf <- st_transform(habspoly_sf, crs=st_crs(coast))
marlpoly_sf <- geojson_sf("data/Entire_marlb_polygons_volume.geojson")
marlpoly_sf <- st_transform(marlpoly_sf, crs=st_crs(coast))

marlpoly_sf.trim <- ms_erase(marlpoly_sf, coast.marlselect) 

globalcrs <- 4326
marlcrs <- st_crs(marlpoly_sf.trim)
```

Locations

```{r}
habloc <- read.csv("data/habloc.csv")
habloc_sf <- st_as_sf(habloc %>% mutate(xp=x,yp=y), 
                      coords=c("xp","yp"),
                      crs=marlcrs,
                      agr="constant")
```

Detections from 1 Jan 2018 onwards

```{r}
event <- read.csv("data/events_alex.csv")
nrow(event) # 6527
event$date <- as.Date(event$date)
event <- event[event$include,]
nrow(event) # 6327

# subset the events
locs <- habloc[habloc$CLUSTER_ID%in%event$CLUSTER_ID,]
locs_sf <- habloc_sf[habloc_sf$CLUSTER_ID%in%event$CLUSTER_ID,]
```

Need one event per polygon per week - just take the first event for simplicity
```{r}
eventpoly <- event %>%
  group_by(CLUSTER_ID,week) %>%
  arrange(date) %>%
  summarise(fevent=first(event)) %>%
  ungroup()
nrow(event) # 6372
nrow(eventpoly) # 6351

event$first <- event$event %in% eventpoly$fevent
table(event$first) # 21  6351
table(event$alex,event$first) # lose 15 detections from this
event <- event[event$first,]
nrow(event) # 6351
```

Find number of detections and max observation value per polygon
```{r}
polydat <- event %>%
  group_by(CLUSTER_ID) %>%
  summarise(lon=first(lon),lat=first(lat),x=first(x),y=first(y),nobs=n(),maxValue=max(Value)) %>%
  ungroup()
#View(polydat)

polydat_sf <- merge(marlpoly_sf.trim, polydat, all.x=TRUE)
polydat_sf$nobs[is.na(polydat_sf$nobs)] <- 0
```


```{r message=FALSE, warning=FALSE}
plot((polydat_sf %>% filter(nobs>0))["maxValue"], key.pos=4, reset=FALSE, 
     main="Maximum Alexandrium detections")
mtext("1-Jan-2018 to 26-Apr-2023", side=1, line=-0.5, cex=0.5, adj=1)#, xlim=xlim.havelock, ylim=ylim.havelock)
plot(st_geometry(coast.marlselect), col="darkgreen", add=TRUE)
plot(st_centroid(st_geometry(marlpoly_sf.trim)), pch="+", col="white", add=TRUE)
plot((polydat_sf %>% filter(nobs>0,maxValue==0))["maxValue"], col="white", add=TRUE)
plot(st_centroid(polydat_sf %>% filter(nobs>0,maxValue==0)), pch="+", col="red", add=TRUE)
```


```{r message=FALSE, warning=FALSE}
plot((polydat_sf %>% filter(nobs>0))["maxValue"], key.pos=4, reset=FALSE, 
     main="Maximum Alexandrium detections - Nydia Bay",
     xlim=xlim.nydia, ylim=ylim.nydia)
mtext("1-Jan-2018 to 26-Apr-2023", side=1, line=-0.5, cex=0.5, adj=1)
plot(st_geometry(coast.marlselect), col="darkgreen", add=TRUE)
plot(st_centroid(st_geometry(marlpoly_sf.trim)), pch="+", col="white", add=TRUE)
plot((polydat_sf %>% filter(nobs>0,maxValue==0))["maxValue"], col="white", add=TRUE)
plot(st_centroid(polydat_sf %>% filter(nobs>0,maxValue==0)), pch="+", col="red", add=TRUE)
```

```{r}
monitored.polygons <- event$CLUSTER_ID
nmonitored <- length(monitored.polygons)
plot(NA,NA, xlim=range(event$date), ylim=c(0,nmonitored+1), xlab="Date", ylab="", axes=FALSE)
axis.Date(1); axis(2, at=1:nmonitored, lab=monitored.polygons, las=2, cex.axis=0.5); box()
idx <- event$CLUSTER_ID%in%monitored.polygons
#points(events.alex.week.2018$mDate[idx], match(events.alex.week.2018$CLUSTER_ID[idx],monitored.polygons),
#       pch=16, cex=2*events.alex.week.2018$Value[idx]/1000/100, col="#00995555")
points(event$date[idx], match(event$CLUSTER_ID[idx],monitored.polygons),
       pch=16, cex=0.5*log(event$Value[idx]/1000), col="#00995555")
title("Detections by polygon over time (log scale)")
```

```{r warning=FALSE, message=FALSE}
wmin <- min(event$week)
wmax <- max(event$week)

par(opar); par(oma=c(3.1,0.5,0,0)); par(mar=1.1*c(2.5,1,4.1,4.1))
logscale <- TRUE
#logscale <- FALSE
if(logscale) {
  xrange <- log(1+range(event$Value)/1000)
} else {
  xrange <- range(event$Value)/1000
}
nbreaks <- 9
breaks <- seq(from=xrange[1], to=xrange[2], length=nbreaks)
colvec <- colorRampPalette(c("light grey","red"))(nbreaks-1)
outdir <- c("ignore/fig1/")
fstem <- "poly1"
ofile <- paste0("ignore/",fstem,".gif")
interactive <- FALSE
#interactive <- TRUE
wrange <- wmin:wmax 
#wrange <- wmin+0:10
xlim <- xlim.marlselect; ylim <- ylim.marlselect
xlim <- xlim.nydia; ylim <- ylim.nydia
#date.axis <- FALSE
date.axis <- TRUE
i <- 0
for(w in wrange) { 
  thisweek.poly <- merge(marlpoly_sf.trim, 
             event[event$week==w,c("CLUSTER_ID","Value")] %>% 
               mutate(Value1000=Value/1000)) 
  if(logscale) thisweek.poly$Value1000 <- log(1+thisweek.poly$Value1000)
  i <- i+1
  fname <- sprintf("%s/%s%04d.png", outdir, fstem, i)
  if(!interactive) png(file=fname, width=480, height=480)
  par(oma=c(3.1,0.5,0,0))
  par(mar=1.1*c(2.5,1,4.1,4.1))
  plot(thisweek.poly["Value1000"], 
       pal=colvec,
       key.pos=4,
       breaks=breaks,
       reset=FALSE, 
       xlim=xlim, ylim=ylim, main="", axes=FALSE)
  plot(st_geometry(coast.marlselect), col="light green", add=TRUE)
  title(main=paste0("Week ",w))
  box()
  mtext(format(week.as.date(w), "%d-%m-%Y"), 
        side=3, line=0, adj=1, cex=0.8)
  mtext(ifelse(logscale,"Log scale","Linear scale"), 
        side=3, line=0, adj=0, cex=0.8)
  if(date.axis) {
    par(usr=c(as.numeric(week.as.date(wrange[c(1,length(wrange))])),0,1))
    axis.Date(1)
    points(week.as.date(w),0,pch=16,cex=2,xpd=TRUE)
  }
  if(!interactive) dev.off()
  #plot(st_centroid(st_geometry(marlpoly_sf.trim)), pch="+", col="white", add=TRUE)
}
par(opar)
if(!interactive) {
  image_list <- lapply(list.files(outdir,full.names=TRUE), image_read)
  poly_animated <- image_animate(image_join(image_list), fps=20)
  image_write(poly_animated, ofile)
  unlink(list.files(outdir,full.names=TRUE))
}
```


# Connectivity and Migration matrices

In the matrices below the rows are the destination locations ($i=1,\dots,n$) 
and the columns are the source locations ($j=1,\ldots,n$).

Notation: at time step $t\rightarrow t+1$ we release Mass $M_{t.j}$ from polygon $j$.  
Mass $M_{tij}$ arrives in polygon $i$ from polygon $j$, thus
\[
   M_{t.j} = \sum_{i=1}^n M_{tij}
\]
The proportion of mass leaving polygon $j$ at time $t$ and arriving at polygon $i$ at time $t+1$ is
\[
   P_{tij} = \frac{M_{tij}}{M_{t.j}}
\]
with row sums
\[
   \sum_{i=1}^n P_{tij} = 1\qquad \text{for all polygons $j$ and times $t$}
\]
$P_t = (P_{tij})$ is the **connectivity matrix** at time $t$.

The total mass arriving in polygon $i$ from all sources (including itself) is 
\[
   M_{ti.} = \sum_{j=1}^n M_{tij}
\]
The proportion of mass arriving in polygon $i$ at time $t+1$ that came from polygon $j$ at time $t$ is
\[
   Q_{tij} = \frac{M_{tij}}{M_{ti.}}
\]
with column sums
\[
   \sum_{i=1}^n Q_{tij} = 1\qquad \text{for all polygons $j$ and times $t$}
\]
$Q_t = (Q_{tij})$ is the **migration matrix** at time $t$.


**Notes from Romain**

I am sending you 2 matrices with two different ways of computing the connectivity. 
Each file has one sheet per release day and each sheet shows the connectivity after 7 days of tracking. 

Rows represent the source locations and columns represent the receiving locations (ordered following the sampling stations ID). 

In Connectivity_matrices_04_to_07_2018, you will find the connectivity represented as a percentage of the 
total released at each sampling station, so each row almost add to 1 (some particles are not within 
a polygon at the end of the run, and therefore they are not counted).

In Migration_matrices_04_to_07_2018, you will find the connectivity normalized
per column (sum to 1 per column).  This gives you the percentage of the contribution 
of each source location to a receiving location.

**RA:** We reverse the rows/columns when reading in: connectivities ($P$) should 
then have unit column sums, and migrations ($Q$) have unit row sums

Note range=cell_cols(1:408) specifies to read 408 columns - otherwise leading blank columns get dropped

```{r}
getmatrices <- TRUE
getmatrices <- FALSE
if(getmatrices) {
  # Connection and migration matrices for the polygons
  fname <- "data/connectivity_matrix.xlsx"
  nsheets <- length(excel_sheets(fname))
  conn <- lapply(1:nsheets, function(i) t(as.matrix(read_excel(fname, sheet=i, 
                                                               col_names=FALSE,
                                                               range=cell_cols(1:408)))))
  names(conn) <- excel_sheets(fname)

  #fname <- "data/Migration_matrices_04_to_07_2018.xls"
  fname <- "data/migration_matrix.xlsx"
  nsheets <- length(excel_sheets(fname))
  migr <- lapply(1:nsheets, function(i) t(as.matrix(read_excel(fname, sheet=i, 
                                                               col_names=FALSE,
                                                               range=cell_cols(1:408)))))
  names(migr) <- excel_sheets(fname)
  matdates <- as.Date(names(migr),format="%Y%m%d")
}
#save.image()
```

```{r}
sapply(conn,dim)-408
sapply(migr,dim)-408
```

Correctly identifies the columns as source polygons, rows as destinations

The trouble is that particles have only been released in 40 of the polygons: not all 408

```{r}
dim(migr[[1]])

range(apply(conn[[1]],1,sum,na.rm=TRUE)) # row sums 0 1.802315
range(apply(conn[[1]],2,sum,na.rm=TRUE)) # col sums 0 1

table(apply(conn[[1]],1,sum,na.rm=TRUE)>0) # 320 nonzero,  88 zero
table(apply(conn[[1]],2,sum,na.rm=TRUE)>0) #  40 nonzero, 368 zero
#  conn[i,j] has sum_i conn[i,j] = 1 for all j where particles have been released
#  i.e. conn[i,j] = proportion of mass from j that ends up in polygon i
#       conn[i,j] = P_{ij}

hist(apply(conn[[1]],1,sum,na.rm=TRUE))
hist(apply(conn[[1]],2,sum,na.rm=TRUE))
```

```{r}
range(apply(migr[[1]],1,sum,na.rm=TRUE)) # row sums 0 1
range(apply(migr[[1]],2,sum,na.rm=TRUE)) # col sums 0 113.2058

table(apply(migr[[1]],1,sum,na.rm=TRUE)>0) # 320 nonzero,  88 zero
table(apply(migr[[1]],2,sum,na.rm=TRUE)>0) #  40 nonzero, 368 zero
# migr[i,j] has sum_j migr[i,j] = 1 for all i where particles have been released
# i.e. migr[i,j] = proportion of mass in i that came from one of the source polygons j

hist(apply(migr[[1]],1,sum,na.rm=TRUE))
hist(apply(migr[[1]],2,sum,na.rm=TRUE))
```


Plot water residency by polygon
```{r}
diag(conn[[1]])
modelled.poly <- which(apply(conn[[1]],2,sum)>0)
notmodelled.poly <- which(apply(conn[[1]],2,sum)==0)
```

```{r}
residency.list <- as.data.frame(sapply(conn, function(pmat) diag(pmat)))
residency.list[notmodelled.poly,] <- NA
residency.list <- data.frame(CLUSTER_ID=1:nrow(residency.list), residency.list)
residency.list[1:3,]
```

```{r}
xrange <- range(residency.list[,-1])
breaks <- seq(from=0, to=1, length=9)
plot(merge(marlpoly_sf.trim, residency.list)["X20180404"],
     key.pos=4, reset=FALSE,
     breaks=breaks, main="Residency")
```

```{r}
xrange <- range(residency.list[,-1])
breaks <- seq(from=0, to=1, length=9)
plot(merge(marlpoly_sf.trim, residency.list)["X20180404"],
     key.pos=4, reset=FALSE,
     breaks=breaks, main="Residency",
     xlim=xlim.nydia, ylim=ylim.nydia)
box()
```


```{r}
print(poly_animated)
```

Which polygons are in the Nydia area?
```{r}
idx <- ((marlpoly_sf.trim$x>xlim.nydia[1] & marlpoly_sf.trim$x<xlim.nydia[2]) & 
        (marlpoly_sf.trim$y>ylim.nydia[1] & marlpoly_sf.trim$y<ylim.nydia[2]) ) 
marlpoly_sf.trim[idx,]
```



