---
title: "HAB model"
date: '2024-04-02'
output:
  pdf_document: 
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: preamble.tex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

<!--- For HTML Only --- Equivalent code for .pdf must be included in preamble.tex -->
`r if (!knitr:::is_latex_output()) '
$\\DeclareMathOperator*{\\argmin}{argmin}$
$\\DeclareMathOperator*{\\logit}{logit}$
$\\newcommand{\\var}{\\mathrm{Var}}$
$\\newcommand{\\bfa}[2]{{\\rm\\bf #1}[#2]}$
$\\newcommand{\\rma}[2]{{\\rm     #1}[#2]}$'`

```{r error=FALSE, warning=FALSE, message=FALSE}
# Packages + set up
load("opar.Rda")
library(readxl)
library(dplyr)
library(tidyverse)
library(tidyr)
library(readxl)
library(ggplot2)
library(maps)
library(sf)
library(geojsonsf)
library(rmapshaper)
library(magick)
library(colorspace)
library(kableExtra)

library(algae)
source("funcs.R")
source("maps.R")
source("setup.R")
```

## Model

 * $n$ polygons $i=1,\ldots,n$
 * $t$ time in weeks: Week 0 is the week of 1 January 2018, $t=0,\ldots,T$
 * $q_t$ is the seasonal indicator: 1 if the week is in the months of December-July, 0 otherwise
 * $V_i$ polygon volume to a depth of 15m
 * $A_i$ polygon area
 * $P_{ijt}$ = proportion of mass in polygon $j$ moving to polygon $i$ in week $t$. 
 * Retention: $W_{jt} \equiv \sum_{i=1}^n P_{ijt}$ proportion of mass in polygon $j$ 
   retained by the system (i.e. not lost at the edges) over the one week period at time $t$
 * Mass conservation implies
   \[ 
     W_{jt} = \sum_{i=1}^n P_{ijt}\leq 1 \qquad\text{for all $j,t$}
   \]
   If the sum is less than 1 when mass is lost at the edges of the set of polygons.
 * Residency: $R_{it} \equiv P_{iit}$ = proportion of mass remaining resident in polygon $i$ in week $t$
 * ${\bf x}_{it}$ covariates in polygon $i$ at time $t$ (temperature, salinity, etc.), excluding residency
   where the sum is less than 1 when mass is lost at the edges of the set of
   polygons.
 * ${\bf x}_{it}$ covariates in polygon $i$ at time $t$ (temperature, salinity, etc.)
 
 * Polygon carrying capacity (density)
   \[
      \rho_{cit} = e^{\alpha_c + {\bf x}_{it}^T\beta_c + \lambda_c\log R_{it}}
   \]
 * Polygon carrying capacity (mass)
   \[
      M_{cit} = V_i\rho_{cit} = V_i e^{\alpha_c + {\bf x}_{it}^T\beta_c + \lambda_c\log R_{it}}
   \]

 * At time step $t$ the epoch indicator $E_t|E_{t-1}$ is drawn
   \begin{eqnarray*}
      E_0           &\sim& \text{Bernoulli}(\mbox{$\frac12$})\\
      E_t|E_{t-1}=0 &\sim& \text{Bernoulli}(\tau_0 q_t)\qquad\text{for $t>0$}\\
      E_t|E_{t-1}=1 &\sim& \text{Bernoulli}(\tau_1 q_t)
   \end{eqnarray*}
   where $\tau_0$ is the probability of initiation of an epoch favourable to algal blooms 
   ($\tau_0$ is low), and $\tau_1$ is the probability of the persistence of that epoch.
 * Growth rate in polygon $i$ at time $t$
   \[
      Q_{it} = e^{ -\gamma_2(1-E_t) 
                          + E_t(\alpha_2 + {\bf x}_{it}^T{\beta}_2 + \lambda_2\log R_{it}) 
      }
   \]
   $\alpha_2>0$ is the basic (covariate independent) growth rate of blooms in a favourable epoch $E_t=1$,
   and $\lambda_2>0$.  
   
   $\gamma_2>0$ is the decay rate of blooms in an unfavourable epoch $E_t=0$.
 * $\pi_{it}$ bloom initiation probability in polygon $i$ at time $t$
   \[
      \log\pi_{it} = \log V_i + \alpha_0 + {\bf x}_{it}^T{\beta}_0 + \lambda_0 \log R_{it}
   \]
   Note that $\pi_{it}<\!\!<1$
 * At time step $t$ in each polygon $i$ the innovation indicator is drawn
   \[
      I_{it}|E_t \sim \text{Bernoulli}(\pi_{it}E_t)
   \]
   An innovation in biomass only occurs in a polygon if $t$ is in a favourable epoch ($E_t\neq 0$).
 * New biomass innovation in polygon $i$ at time $t$
   \begin{eqnarray*}
      B_{it} | I_{it}=0 &\sim& \delta_0\\
      B_{it} | I_{it}=1 &\sim& \text{Gamma}(a,b)
   \end{eqnarray*}
   $\delta_0$ is a point mass at zero: we get no new biomass if $I_{it}=0$.
 * Baseline algal mass distribution $M_{i0}=0$ for all $i$: i.e. no mass at time $t=0$.
 * Time evolution of algal mass $M_{it}$ for $t>0$
   \begin{eqnarray*}
      D_{it} &=& B_{it} + \sum_{j=1}^n P_{ijt} e^{E_t({\bf x}_{jt}^T{\beta}_1 + \lambda_1 \log R_{jt})} M_{j,t-1}\\
      M_{it} &=& e^{\eta_{it}} M_{cit} \left[1 - e^{-Q_{it}D_{it}/M_{cit}}\right]
   \end{eqnarray*}
   where
   \[
      \eta_{it} \sim \text{Normal}(0,\sigma^2_\eta)
   \]
   We may not need both of $({\lambda_1,\beta}_1)$ (growth before transport) in addition to 
   $(\lambda_2,{\beta}_2)$ (growth after transport, within $Q_{it}$).
 * Observation of algal density $Y_{it}$
   \begin{eqnarray*}
     Y_{it} | M_{it} = 0 &\sim& \delta_0\\
     \log Y_{it} | M_{it}>0, \sigma_\varepsilon^2 &\sim& \text{Normal}(\log (M_{it}/V_i),
     \sigma_\varepsilon^2)
   \end{eqnarray*}


Summary of quantities

Indexing

* $i=1,\ldots,n$ - labels of polygons
* $t=0,\ldots,T$ - labels of weeks

Fixed

* $q_t$ seasonal indicator (1 if the week is in the months of December-July, 0 otherwise)
* $V_i$ volume of polygon $i$ to a depth of 15m
* $A_i$ area of polygon $i$

Observed/Modelled separately

* ${\bf x}_{it}$ - covariates (temperature, salinity, etc)
* $P_{ijt}$ - proportion of mass in polygon $j$ moving to polygon $i$ in week $t$
* $Y_{it}$ - observed density (in a subset of polygons at a subset of times)

Latent (to be estimated)

* $E_t$ epoch indicator
* $I_{it}$ innovation indicator
* $B_{it}$ innovation biomass (where $I_{it}=1$)
* $M_{it}$ mass in polygon $i$ at time $t$

Derived/Modelled

* $R_{it}$ - residency: derived from ${\bf P}_t$: $R_{it} = P_{iit}$
* $\pi_{it}$ - bloom initiation probability: $\log\pi_{it}=\log V_i + \alpha_0 + {\bf x}_{it}^T{\beta}_0+\lambda_0\log R_{it}$
* $M_{cit}$ - carrying capacity: $\log M_{cit}=\log V_i + \alpha_c + {\bf x}_{it}^T{\beta}_c + \lambda_c\log R_{it}$
* $Q_{it}$ - growth rate: $\log Q_{cit}=-\gamma_2(1-E_t) + E_t(\alpha_2 + {\bf x}_{it}^T{\beta}_2 + \lambda_2\log R_{it})$
* $D_{it}$ - transported mass: $D_{it}=B_{it} + \sum_{j=1}^n P_{ijt} e^{{\bf x}_{it}^T{\beta}_1 + \lambda_1\log R_{it}} M_{j,t-1}$

Parameters (to be estimated)

* $\tau_0$ growth epoch initiation probability
* $\tau_1$ growth epoch persistence probability
* $(\gamma_2, \alpha_2)$ baseline log growth parameters in epochs 0 (decay) and 1 (growth)
* $(\alpha_c,\lambda_c,\beta_c)$ parameters of carrying capacity
* $(\alpha_0,\lambda_0,\beta_0)$ parameters of bloom initiation probability
* $(\lambda_1,\beta_1)$ parameters of pre-transport growth
* $(\lambda_2,\beta_2)$ parameters of post-transport growth
* $\sigma_\eta$ - mass evolution error
* $\sigma_\varepsilon$ - density observation error

## Simplest model

 * Carrying capacity of polygon $i$ at time $t$
   \[
      M_{cit} = V_i \rho_{cit} = V_i \rho_c R_{it}^{\lambda_c} = V_i e^{\alpha_c + \lambda_c \log R_{it}}
   \]
 * $\pi_{it}$ bloom initiation probability in polygon $i$ at time $t$
   \[
      \log\pi_{it}  = \log V_i + \alpha_0 + \lambda_0 \log R_{it}
   \]
 * Epoch
   \begin{eqnarray*}
      E_0           &\sim& \text{Bernoulli}(\mbox{$\frac12$})\\
      E_t|E_{t-1}=0 &\sim& \text{Bernoulli}(\tau_0 q_t)\qquad\text{for $t>0$}\\
      E_t|E_{t-1}=1 &\sim& \text{Bernoulli}(\tau_1 q_t)
   \end{eqnarray*}
 * Growth rate in polygon $i$ at time $t$
   \[
      Q_{it} = e^{ -\gamma_2(1-E_t) 
                          + E_t(\alpha_2 + {\bf x}_{it}^T{\beta}_2 + \lambda_2\log R_{it}) 
      }
   \]
 * At time step $t$ in each polygon $i$ the innovation indicator is drawn
   \[
      I_{it}|E_t \sim \text{Bernoulli}(\pi_{it}E_t)
   \]
   An innovation in biomass only occurs in a polygon if $t$ is in a favourable epoch ($E_t\neq 0$).
 * New biomass innovation in polygon $i$ at time $t$
   \begin{eqnarray*}
      B_{it} | I_{it}=0 &\sim& \delta_0\\
      B_{it} | I_{it}=1 &\sim& \text{Gamma}(a,b)
   \end{eqnarray*}
 * Baseline algal mass distribution $M_{i0}=0$ for all $i$: i.e. no mass at time $t=0$.
 * Time evolution of algal mass $M_{it}$ for $t>0$
 
   Transport: 
   \begin{eqnarray*}
      D_{it} &=& B_{it} + \sum_{j=1}^n P_{ijt} M_{j,t-1}
   \end{eqnarray*}
   Growth
   \begin{eqnarray*}
      M_{it} &=& e^{\eta_{it}} M_{cit} \left[1 - e^{-Q_{it}D_{it}/M_{cit}}\right]
   \end{eqnarray*}
   where
   \[
      \eta_{it} \sim \text{Normal}(0,\sigma^2_\eta)
   \]
 * Observation of algal density $Y_{it}$
   \begin{eqnarray*}
     Y_{it} | M_{it} = 0 &\sim& \delta_0\\
     \log Y_{it} | M_{it}>0, \sigma_\varepsilon^2 &\sim& \text{Normal}(\log (M_{it}/V_i),
     \sigma_\varepsilon^2)
   \end{eqnarray*}
   ... add a possibility of false negatives ... at each station?  or a single 
   probability for all stations?


## Data in the Nydia area

```{r, warning=FALSE}
nydia_sf <- st_crop(marlpoly_sf.trim, 
                    xmin=xlim.nydia[1], xmax=xlim.nydia[2], ymin=ylim.nydia[1], ymax=ylim.nydia[2])
nydia_sf <- nydia_sf %>% filter(!(CLUSTER_ID%in%c(48,71,162,194,251,258,293,406)))
nydia.poly <- nydia_sf$CLUSTER_ID

idx <- nydia.poly
wvec <- date.as.week(matdates)
connsub <- lapply(conn, function(cmat) cmat[idx,][,idx])
ressub <- sapply(connsub,diag)
covsublist <- lapply(covlist, function(cmat) cmat[idx,])
eventsub <- event[event$CLUSTER_ID%in%idx,]
#nrow(eventsub) # 541
#length(unique(eventsub$week)) # 274
#length(unique(eventsub$CLUSTER_ID)) # 11 monitored polygons
#unique(eventsub$CLUSTER_ID)

eventsubw <- eventsub[eventsub$week%in%wvec,]
#nrow(eventsubw) # 12
#length(unique(eventsubw$week)) # 12
#length(unique(eventsubw$CLUSTER_ID)) # 1 # just one polygon: 400

pmat <- t(connsub[[1]])
#dim(pmat) # 17 17
n <- nrow(pmat) # 17 polygons
nydia_sf$nid <- 1:n
#hist(apply(conn[[1]],2,sum))
resvec <- diag(pmat)
```

```{r warning=FALSE}
plot(nydia_sf %>% st_geometry())
text(nydia_sf %>% st_centroid(of_largest_polygon=TRUE) %>% st_coordinates(), 
     lab=as.character(nydia_sf$CLUSTER_ID), cex=0.5)
plot(nydia_sf['volume_15'], main="Volume to 15m (cubic km)")
plot((nydia_sf %>% mutate(residency=resvec[nid]))['residency'], main="1 week water residency")
```


```{r}
hist(apply(pmat,1,sum), xlab=expression(W[j]=="Retention "*Sigma[i]*P[ij]),
     main="Retention", breaks=11)
```

## Data on events

```{r}
# Events
# ensure each event has a preceding event, or insert one which has an NA value
plot(NA,NA,xlim=range(eventsub$week),ylim=range(log(1+eventsub$Value)),xlab="Week",ylab="log(1+Value)",
     main="Concentration (log scale)")
invisible(by(eventsub, eventsub$CLUSTER_ID, 
             function(emat) {
                 extra <- (emat$week-1)[!((emat$week-1) %in% emat$week)]
                 extra <- extra[extra>=min(emat$week)]
                 if(length(extra)>0) {
                   emat <- rbind(emat[,c("week","Value")],data.frame(week=extra,Value=NA))
                 } 
                 emat <- emat[order(emat$week),]
                 #print(emat[,c("week","Value")])
                 lines(emat$week,log(1+emat$Value))
             }))
abline(v=wvec[c(1,length(wvec))],col="red")
```

Note that concentration growth is sharp, but then becomes approximately constant.  This is
what motivates the model structure where we have a polygon carrying capacity.  

### Covariates

Residency: we have matrices $P_{ijt}$ for the weeks 2518-2529 (dates `r week.as.date(2518)`-`r week.as.date(2529)`)

```{r}
# Residency
plot(NA,NA,xlim=range(wvec),ylim=c(0,1),xlab="Week",ylab="Residency",main="Residency over time")
invisible(apply(ressub, 1, function(resvec) lines(wvec,resvec)))
axis(4, at=ressub[,ncol(ressub)], lab=nydia.poly, las=2, cex.axis=0.6)
```

Residency is approximately constant during the periods Week 2518-2529.

```{r}
# Events
plot(NA,NA,xlim=range(wvec),ylim=range(log(1+event$Value)),xlab="Week",ylab="log(1+Value)",
     main="Concentration (log scale)")
invisible(by(eventsub, eventsub$CLUSTER_ID, function(emat) lines(emat$week,log(1+emat$Value))))
```

During a bloom concentration rapidly rises, but is then roughly constant.

```{r}
# Temperature
covname <- "Temperature"
xmat <- covsublist[[covname]]
plot(NA,NA,xlim=range(wvec),ylim=range(xmat),xlab="Week",ylab=covname,main=paste(covname,"over time"))
invisible(apply(xmat, 1, function(xvec) lines(wvec,xvec)))
axis(4, at=xmat[,ncol(xmat)], lab=nydia.poly, las=2, cex.axis=0.6)

# Salinity
covname <- "Salinity"
xmat <- covsublist[[covname]]
plot(NA,NA,xlim=range(wvec),ylim=range(xmat),xlab="Week",ylab=covname,main=paste(covname,"over time"))
invisible(apply(xmat, 1, function(xvec) lines(wvec,xvec)))
axis(4, at=xmat[,ncol(xmat)], lab=nydia.poly, las=2, cex.axis=0.6)
```

Temperature over time

```{r}
plot(NA,NA,
     xlim=range(tempdata$date),
     ylim=range(tempdata$tmin,na.rm=TRUE),
     axes=FALSE,xlab="Date",ylab="Daily minimum temperature")
axis.Date(1); axis(2); box()
invisible(by(tempdata,tempdata$Station,function(tt) lines(tt$date,tt$tmin)))
```

```{r}
plot(NA,NA,
     xlim=range(event$date),
     ylim=range(tempdata$tmin,na.rm=TRUE),
     axes=FALSE,xlab="Date",ylab="Daily minimum temperature")
axis.Date(1); axis(2); box()
invisible(by(tempdata,tempdata$Station,function(tt) lines(tt$date,tt$tmin)))
```

## Model implementation

```{r error=FALSE, warning=FALSE, message=FALSE}
# Packages + set up
load("opar.Rda")
library(readxl)
library(dplyr)
library(tidyverse)
library(tidyr)
library(readxl)
library(ggplot2)
library(maps)
library(sf)
library(geojsonsf)
library(rmapshaper)
library(magick)
library(colorspace)
library(kableExtra)

library(algae)
source("funcs.R")
source("maps.R")
source("setup.R")
```

Requirements

* Polygon -sf object, with $n$ polygons
  * `volume_15` column
* Communication matrix $P_{ijt}$ (movement from polygon $j$ to polygon $i$ at time $t$)
* Covariate matrix $X_{it\ell}$ (polygon $i$, time $t$, covariate $\ell$)

## Nydia

```{r}
region <- "nydia"
movie.clean <- FALSE
invisible()
```

```{r child="habmodel-sub1.Rmd"}
```

## Marlborough

```{r}
region <- "all"
movie.clean <- FALSE
invisible()
```

```{r child="habmodel-sub1.Rmd"}
```










